%%=====================================================================
%% List of abbreviations
%%=====================================================================

\begin{listofabbreviations}

\loxline{ACID}{Properties that are required of transactions: atomicity,
consistency, isolation, and durability}{def:acid}

\loxline{CMVBT}{Concurrent multiversion \Btree. 
A fully concurrent index structure composed of a TMVBT index and a
main-memory-resident VBT index}{def:cmvbt}

\loxline{CTI}{Commit-time-to-identifier table.  
This table is used in the CMVBT index to map commit-time versions into
transaction identifiers}{def:cti}
 
\loxline{DAG}{Directed acyclic graph}{def:dag}
 
\loxline{Immortal~DB}{A multiversion database system built on top of
Microsoft SQL Server. 
Based on the \TSBtree\ index structure}{def:immortaldb} 

\loxline{LRU}{Least recently used}{def:lru}

\loxline{LSN}{Log sequence number}{def:lsn}

\loxline{MBR}{Minimum bounding region}{def:mbr}

\loxline{MVAS}{Multiversion access structure.
A multiversion index structure designed by Varman and Verma. 
Similar to the MVBT}{def:mvas}

\loxline{MVBT}{Multiversion \Btree. An asymptotically optimal multiversion
index by Becker et~al.}{def:mvbt}

\loxline{PTT}{Persistent timestamp table. 
The PTT is used to map transaction identifiers into
commit-time versions in the \TSBtree. 
It is similar to the CTI table but has a reverse mapping}{def:ptt}
 
\loxline{SI}{Snapshot isolation, a commonly used multiversion
concurrency-control algorithm that can be made serializable}{def:si}

\loxline{SMO}{A structure-modification operation is 
an operation that modifies the structure of the database index by
splitting or merging a database page}{def:smo}
 
\loxline{TMVBT}{Transactional multiversion \Btree. 
Our transactional extension to the MVBT index. 
Allows a single updating transaction to run concurrently with
multiple read-only transactions}{def:tmvbt}

\loxline{\TSBtree}{Time-split \Btree. 
A multiversion index structure by Lomet and Salzberg}{def:tsb}
 
\loxline{VBT}{Versioned \Btree. 
A \Btree\ used for storing entries that contain version information in
addition to the data item key and value}{def:vbt}

\loxline{VTT}{Volatile timestamp table. Used as a main-memory cache for the
PTT in the \TSBtree}{def:vtt}

\loxline{WAL}{Write-ahead logging. 
A policy that states that the database log file must be flushed onto disk
before any pages that contain the corresponding updates described in
the log are written onto disk}{def:wal}
 
\loxline{WORM}{Write-once, read-many. 
Describes media that can be written to only once, but read from many times,
such as CD-R, DVD-R, BD-R, and tape archives}{def:worm}

\loxline{WMRM}{Write-many, read-many. Describes media that can be written to
multiple times, such as hard drives. 
While this definition includes re-writable discs (CD-RW, DVD-RW, BD-RW), it
is often assumed that WMRM media is faster than WORM media}{def:wmrm}

\end{listofabbreviations}