%%=====================================================================
%% Preface (description of writing the dissertation and thanks)
%%=====================================================================

\begin{preface}

First of all, I would like to thank my supervisor, Professor Eljas
Soisalon-Soininen, for introducing me to the field of algorithm research, and
for providing me with an interesting and challenging research topic.
I have often heard that the topic for a dissertation is half the work; in my
case, I certainly got off to a swift start thanks to being able to start my
research work directly.
I would also like to thank my instructor, Professor Seppo Sippu from the
University of Helsinki, for taking such an active role in supporting my
research work, and for always having time to provide me with exhaustive
comments on my contributions.

I would especially like to thank Professor Peter Widmayer (ETH Zürich) for 
honoring me by agreeing to be the opponent for this dissertation, and I am 
also grateful to my pre-examiners, 
Assistant Professor Dr.~Alexander Markowetz (Rheinische
Friedrich-Wilhelms-Universität Bonn) and
Dr.~Jan Lindström (IBM Helsinki), for their invaluable comments on the
manuscript.
Many interesting future research ideas have already emerged from their
observations. 

My research work was carried out at the Department of Computer Science and
Engineering, and I want to thank all my colleagues for their support and
company.
I thank my team colleague, Dr.~Ibrahim Jaluta, for introducing me to the
myriad details of database algorithms and for always providing me with timely
research papers on the research topics at hand.
My work and post-graduate studies have been funded by the Helsinki Graduate
School in Computer Science and Engineering, and by the Academy of Finland,
for which I am grateful.

I thank my friends and family for their support and encouragement; I would
especially like to thank my mother for setting an example I could aspire to.
Finally, I thank my darling wife Anu for her love and support, for her
patience in having to listen to my rants on database research, and
also for her practical help with the dissertation. 

\begin{flushright}
Espoo, September 2010\\
\vspace{2ex}
Tuukka Haapasalo
\end{flushright}

\end{preface}